Spfy: distributed predictive genomics of E.coli with graph based result linkage

### 2017-08-25 paper-webserver.tex
ADD: one line
ADD: general edits

\section{FUNCTIONALITY}
% ONLY FOCUS ON THE ANALYSIS MODULES
% Describe available functions in spfy
% para covering everything
Spfy performs reference laboratory tasks: O-antigen and H-antigen typing, Shiga-toxin (STX) typing,  virulence factor (VF) and antimicrobial resistance (AMR) gene determination, and related strain determination.
Spfy also performs bioinformatics analyses: pangenome generation, statistical significance testing of genome markers for user-defined groups, and AMR predictions using support vector machines (SVMs).

% para covering ectyper & RGI
To further characterize strains ... serotyping, which detects the presence of specific cell surface antigens.
O-antigen and H-antigen typing, and VF determination use our ECTyper \url{https://github.com/phac-nml/ecoli_serotyping} software.
ECTYper compares the user-submitted genomes against a curated set of sequences, an approach originally implemented in the VirulenceFinder \cite{joensen2014real} tool.
The Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} maintains a set of known AMR genes and develops the Resistance Gene Identifier (RGI) program.
Both ECTyper and RGI are based off of BLAST \cite{pmid2231712}.

@@

\subsection{Data Storage}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries
Graph databases focus on describing relationships between different data points, is one of the emerging \cite{de2015trends} database types used for biological data, and one of the core building blocks of a semantic web technology stack \cite{horrocks2005semantic}.
Spfy uses the RDFLib Python library \url{https://rdflib.readthedocs.org/} to represent all data meant for long-term storage.
When serotyping, VF, AMR predictions, and pangenome generation tasks are completed, the results are stored within the graph database.
% note: below is a feature in development
The permanent storage of results serve as a one-time cost, and allows population-wide analyses of all stored genomes; result storage also enables Spfy to avoid recomputation when the same analysis is re-run.

@@

The front-end website displayed to users is written using the React JavaScript library \url{https://facebook.github.io/react/} as a single-page application to allow efficient data-flow without reloading the website.
To ensure a familiar user interface, we followed the Material Design specification \url{https://material.io/}, published by Google, and follow a card-based design.
(see Figure \ref{fig-results})
While data storage is graph-based, the results of various analysis modules are presented to users in a familiar tabular structure and available for download.
(see Figure \ref{fig-tables})

@@

Task queues are processes which schedule code exection across available computing infrastructure.
In Spfy, the Python-based Redis Queue library \url{https://github.com/nvie/rq} is used to schedule analysis tasks and have them run asynchronously in response to user requests.
When a user submits files for analysis or population-wide anlayses, separate tasks are enqueued at different priorities, depending on when users might expect the result to be returned.
For example, population-wide analyses have a higher priority than bulk subtyping analyses, as we expect day-to-day web searches (for example, on Google) to respond instantly whereas a delay in subtyping is typical in similar web services such as the Center for Genomic Epidemiology Pipeline (CGE Pipeline) \cite{thomsen2016bacterial}.
Spfy enables processing of thousands of genome sequences by using task queue workers running in parallel, which also allows performance to scale to available infrastructure.
To ensure reliability of the platform, the open-source Sentry toolkit \url{https://github.com/getsentry/sentry} was integrated and used for real-time exception tracking.

@@

TravisCI \url{travisci.io}, a continuous integration (CI) platform, is integrated into Spfy's Github repository \url{https://github.com/superphy/backend} and runs tests ensuring functionality and backwards compatibility with any changes to the codebase.
The individual tests use PyTest \url{https://doc.pytest.org/} and are run within TravisCI's virtual environment.
The current build status can be checked either on our Gtihub repository or at \url{https://travis-ci.org/superphy/backend}.

@@

% e.g.
% database statistics
We tested spfy with 55,353 samples genomes (267GB), storing both the entire genomes and results for all included analyses modules.
The resulting database had XYZ nodes and XYZ edges, with XYZ object properties.
This worked our to XYZ GB of data stored.

@@

% namely galaxy
Other scientific workflow technologies such as Galaxy \cite{goecks2010galaxy}, Kepler \cite{ludascher2006scientific}, and Taverna \cite{oinn2004taverna} have been applied to bioinformatic tasks.
Galaxy ...
Taverna ...
Kepler ...
To approach the challenge of integrating different bioinformatics programs, Spfy instead uses technologies prevalent in common web services not necessarily related to scientific workflows.
While we prefer to package individual bioinformatics programs using language-agnostic package managers such as Conda \url{https://conda.io/}, we do agree that suites of bioinformatics software benefit from having all dependencies installed and deployable in Docker containers \cite{di2015impact}.
These programs are connected to form a workflow using task queues; a design methodology used to offload long-running or non-instantaneous tasks over available computing infrastructure.
Unlike platforms such as Galaxy which favor a drag-and-drop design for building workflows, Spfy favors a tightly coupled approach explicitedy linking together different bioinformatics programs within the task queue implemntation.
In particular, Spfy uses asynchronous tasks queues which are tripped in response to user requests, thereby allowing immediate responsivity to users on the website, and processes generating human-readable results can be separated from processes handling long-term result storage.

% standard web tech means you can deploy to different cloud computing services
One of the key benefits of using more common-place technologies is the compatibility with other infrastructure resources.
Docker containerization is widely supported by cloud computing services: Amazon Web Services (AWS) \url{https://aws.amazon.com/docker/}, Google Cloud Platform (GCloud) \url{https://cloud.google.com/container-engine/}, and Microsoft Azure \url{https://azure.microsoft.com/en-us/services/container-service/}, and self-hosted cloud computing technologies such as OpenStack \url{https://wiki.openstack.org/wiki/Docker}.
% we may want to sign up for a free trial to give an example of this
Spfy packages compute nodes, which manage a collection of task queue workers, as reproducible Docker containers which can be networked together to form a compute cluster.
Docker containerization has a neglible impact on performance \cite{di2015impact}, and allows the platform to easily scale to demand.
In contrast to Galaxy, Spfy is a targeted approach building a predictive analytics platform for \textit{E. coli} which leverages generic web technologies to favor a responsive, big-data design strategy tackling the challenge \cite{fricke2014bacterial} of integrating results from multiple genomes and examining shared connections.

@@

% para: \cite{joensen2014real,thomsen2016bacterial}
% we have a lot of similar directions to this, but just try to make it prettier, more friendly to use, and faster
The Bacterial Analysis Platform (BAP), developed out of the Technical University of Denmark, provides and integrated analysis pipeline for bacterial WGS data, as a web service.
BAP is novel in its combined approach to genome analysis as different programs, such as for VF and AMR determination, are included by default into the pipeline; Spfy provides similar functionallity with an expanded focus on integrated result storage and big-data analyses.
In place of a MySQL database for storing the location of result files, Spfy parses result data into graph objects and integrates results into a persistant graph database.
This enables Spfy to create large datasets where analysis results are linked to genomes, and to perform population-wide analyses.
On a per file basis, Spfy performs at a similar speed to BAP on predictive genomics tasks, though Spfy does not provide genome assembly services.
Spfy also processes XXXX files over XX tasks in XXXX time by using a novel approach of distributing computations over a task queue and multiple Docker compartmentalized containers.

>>>>

\section{FUNCTIONALITY}
% ONLY FOCUS ON THE ANALYSIS MODULES
% Describe available functions in spfy
% para covering everything
Spfy performs reference laboratory tasks: O-antigen and H-antigen typing, Shiga-toxin (STX) typing,  virulence factor (VF) and antimicrobial resistance (AMR) gene determination, and related strain determination.
Spfy also performs bioinformatics analyses: pangenome generation, statistical significance testing of genome markers for user-defined groups, and AMR predictions using support vector machines (SVMs).
We introduce one new bioinformatics program, Ecoli-Serotyper (ECTyper) \url{https://github.com/phac-nml/ecoli_serotyping}, for determining surface-antigen type and associated VFs, and utilize existing software: the Resistance Gene Identifier (RGI) program \cite{mcarthur2013comprehensive}, Phylotyper \cite{whiteside2017phylotyper}, and Panseq \cite{laing2010pan}.

% para covering ectyper & RGI
To further characterize strains ... serotyping, which detects the presence of specific cell surface antigens.
O-antigen and H-antigen typing, and VF determination use our ECTyper software.
ECTYper compares the user-submitted genomes against a curated set of sequences, an approach originally implemented in the VirulenceFinder \cite{joensen2014real} tool.
The Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} maintains a set of known AMR genes and develops the RGI program .
Both ECTyper and RGI are based off of BLAST \cite{pmid2231712}.

@@

\subsection{Data Storage}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries
Graph databases describe relationships between different data points, is one of the emerging \cite{de2015trends} database types used for biological data, and is one of the core building blocks of a semantic web technology stack \cite{horrocks2005semantic}.
Spfy uses the RDFLib Python library \url{https://rdflib.readthedocs.org/} to represent all data meant for long-term storage.
When serotyping, VF, AMR predictions, and pangenome generation tasks are completed, the results are stored within the graph database.
% note: below is a feature in development
The permanent storage of results serve as a one-time cost, and allows population-wide analyses of all stored genomes; result storage also enables Spfy to avoid recomputation when the same analysis is re-run.

@@

The front-end website is written using the React JavaScript library \url{https://facebook.github.io/react/} as a single-page application to allow efficient data-flow without reloading the website.
To ensure a familiar user interface, we followed the Material Design specification \url{https://material.io/}, published by Google, surrounding a card-based design.
(see Figure \ref{fig-results})
Both the task selection and results display follow this card-based design: while data storage is actually graph-based, the results of various analysis modules are presented to users in a familiar tabular structure and available for download.
(see Figure \ref{fig-tables})

@@

Task queues are processes which schedule code exection across available computing infrastructure.
In Spfy, the Python-based Redis Queue library \url{https://github.com/nvie/rq} is used to manage analysis tasks and have them run asynchronously in response to user requests.
When a user submits files for analysis or population-wide anlayses, separate tasks are enqueued at different priorities, depending on when users might expect the result to be returned.
For example, population-wide analyses have a higher priority than bulk subtyping analyses, as we want search-response queries to respond instantly whereas a delay in subtyping is typical in similar web services such as the Center for Genomic Epidemiology Pipeline (CGE Pipeline) \cite{thomsen2016bacterial}.
Spfy enables processing of thousands of genome sequences by using task queue workers running in parallel, which also allows performance to scale to available infrastructure.
To ensure reliability of the platform, the open-source Sentry toolkit \url{https://github.com/getsentry/sentry} was integrated and used for real-time exception tracking.

@@

TravisCI \url{travisci.io}, a continuous integration (CI) platform, is integrated into Spfy's Github repository \url{https://github.com/superphy/backend} and, with any changes to the codebase, runs tests for functionality and backwards compatibility.
The individual tests use PyTest \url{https://doc.pytest.org/}, are run within TravisCI's virtual environment, and the current build status can be checked either on our Gtihub repository or at \url{https://travis-ci.org/superphy/backend}.
CI is also used to automatically build Spfy's core Docker images, and upload them to Docker Hub \url{https://hub.docker.com/u/superphy/}.

@@

Spfy was tested with 25,185 public \textit{E. coli} genomes, 5,353 genomes from GenBank and 19,832 genomes from Enterobase, and 55,353 generated samples genomes (267GB), storing both the entire sequences and results for all included analysis modules.
The resulting database had XYZ nodes and XYZ edges, with XYZ object properties, which worked out to XYZ GB of data stored.

@@

% namely galaxy
Other scientific workflow technologies such as Galaxy \cite{goecks2010galaxy}, Kepler \cite{ludascher2006scientific}, and Taverna \cite{oinn2004taverna} have been applied to bioinformatic tasks.
Galaxy aims to provide a reproducible compuation-based research environment which is accessible to individuals without programming knowledge.
Taverna ...
Kepler ...
To approach the challenge of integrating different bioinformatics programs, Spfy instead uses technologies prevalent in common web services not necessarily related to scientific workflows.
While we prefer to package individual bioinformatics programs using language-agnostic package managers such as Conda \url{https://conda.io/}, we do agree that suites of bioinformatics software benefit from having all dependencies installed and deployable in Docker containers \cite{di2015impact}.
These programs are connected to form a workflow using task queues; a design methodology used to offload long-running or non-instantaneous tasks over available computing infrastructure.
Unlike platforms such as Galaxy which favor a drag-and-drop design for building workflows, Spfy favors a tightly coupled approach explicitedy linking together different bioinformatics programs within the task queue implemntation.
In particular, Spfy uses asynchronous tasks queues which are tripped in response to user requests, thereby allowing immediate responsivity to users on the website, and processes generating human-readable results can be separated from processes handling long-term result storage.

% standard web tech means you can deploy to different cloud computing services
One of the key benefits of using more common-place technologies is the compatibility with other infrastructure resources.
Docker containerization is widely supported by cloud computing services: Amazon Web Services (AWS) \url{https://aws.amazon.com/docker/}, Google Cloud Platform (GCloud) \url{https://cloud.google.com/container-engine/}, and Microsoft Azure \url{https://azure.microsoft.com/en-us/services/container-service/}, and self-hosted cloud computing technologies such as OpenStack \url{https://wiki.openstack.org/wiki/Docker}.
% we may want to sign up for a free trial to give an example of this
Spfy packages compute nodes, which manage a collection of task queue workers, as reproducible Docker containers which can be networked together to form a compute cluster.
Docker containerization has a neglible impact on performance \cite{di2015impact}, and allows the platform to easily scale to demand.
In contrast to Galaxy, Spfy is a targeted approach building a predictive analytics platform for \textit{E. coli} which leverages generic web technologies to favor a responsive, big-data design strategy tackling the challenge \cite{fricke2014bacterial} of integrating results from multiple genomes and examining shared connections.
By packaging the individual bioinformatics software as Conda packages, they can be easily ported to Galaxy.

@@

% para: \cite{joensen2014real,thomsen2016bacterial}
% we have a lot of similar directions to this, but just try to make it prettier, more friendly to use, and faster
The Bacterial Analysis Platform (BAP), developed out of the Technical University of Denmark, is the closest analogue to Spfy, and provides an integrated analysis pipeline for bacterial WGS data as a web service.
BAP is novel in its combined approach to genome analysis as different programs, such as for VF and AMR determination, are included by default into the pipeline; Spfy provides similar functionallity with an expanded focus on integrated result storage and big-data analyses.
In place of a MySQL database for storing the location of result files, Spfy parses result data into graph objects and integrates results into a persistant graph database.
This enables Spfy to create large datasets where analysis results are linked to genomes, and to perform population-wide analyses.
On a per file basis, Spfy performs at a similar speed to BAP on predictive genomics tasks, though Spfy does not provide genome assembly services.
Spfy also processes XXXX files over XX tasks in XXXX time by using a novel approach of distributing computations over a task queue and multiple Docker compartmentalized containers.
